Making Q-learning robust to time discretization

### 2019-01-17 paper/icml_drau/advup.tex
.

%! TEX root = icml_drau.tex
\section{Reinforcement learning with a continuous time limit}
As seen in Sec.~\ref{sec:framework}, there is no continuous time limit to
Q-learning. In continuous time, $Q^\pi$ is independant of actions and can thus
not be used to select promising actions.  In near continuous time, $Q^\pi_\deltat$
still depends on actions, and can still be used to choose promising actions
under the current policy. However, when approximating $Q^\pi_\deltat$, if the
approximation error is much bigger than $\bigO(\deltat)$, approximation errors
dominates, and the hierarchy of actions given by the approximation is likely to
be erroneous.  When the approximate Q-function is initialized, if the effect of
actions on the Q-function is order of magnitudes higher than what it should be,
approximating $Q^\pi_\deltat$ is likely to be difficult. Besides, the error is likely
to be further propagated when the equation used to update our approximate $Q$
function relies on bootstraping, as is the case for usual temporal difference
derived methods.

In what follows, a discrete algorithm that has a well defined time continuous
limit is defined. The algorithm relies on three elements, namely, defining and
learning a quantity that contains some information on the hierarchy of actions
while admitting a time continuous limit, defining exploration methods that
admit a continuous time limit and defining learning rates that induce well
behaved parameter trajectories when $\deltat$ goes to $0$.

\subsection{Advantage Updating}
\label{subsec:reparam}
To define an object which contains the same information on actions as
$Q^\pi_\deltat$, but admits an action dependant continuous time limit, it is
natural to define
\begin{align}
	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat},
    \label{eq:adv}
\end{align}
a rescaled version of the advantage function, as the difference between between
$Q^\pi_\deltat(s, a)$ and $V^\pi\deltat(s)$ is of order $\bigO(\deltat)$.
Contrary to the action state value function, the rescaled advantage function converges when $\deltat$ goes to $0$
to an action dependent quantity which contains the necessary information for policy improvement:

@@

The discretized Q-function can be rewritten as
\begin{equation}
	Q^\pi_\deltat(s, a) = V^\pi_\deltat(s) + \deltat A^\pi_\deltat(s, a).
	\label{eq:reparam_q_pi}
\end{equation}
A natural way of approximating $V^\pi_\deltat$ and $A^\pi_\deltat$ is to apply
\emph{Sarsa} or Q-learning to a reparameterized Q-function approximator
\begin{equation}
	Q_\Theta(s, a) = V_{\theta}(s) + \deltat A_{\psi}(s, a).
\end{equation}
To avoid cumbersome notations, in what follows, parameter indices are dropped.
At initialization, if both $V_{\theta}$ and $A_{\psi}$ are initialized
independantly of $\deltat$, this parameterization reasonably scales the
contribution of actions relative to states in the value of $Q$.

\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)} However, the proposed reparameterization does not, on its own, guarantee that
when $Q$ correctly approximates $Q^\pi_\deltat$, $A$ approximates $A^\pi_\deltat$.
Indeed, for any given pair $V$, $A$,
\begin{align}
	\tilde{V}(s) &= V(s) - f(s)\\
	\tilde{A}(s, a) &= A(s, a) + \frac{f(s)}{\deltat}
\end{align}
yields the exact same $Q$ function. Consequently, $Q$ can correctly approximate
$Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
To enforce, identifiability of $A$, one must enforce the consistency equation
\begin{equation}
	Q^\pi_\deltat(s, \pi(s)) - V^\pi_\deltat(s) = 0
\end{equation}
on the approximate $A$ and $V$. This translates to
\begin{equation}
	A(s, \pi(s)) = 0.
\end{equation}
With this additional constraint, if $Q^\pi_\deltat = Q$,
\begin{align}
	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}\\
		    &= A(s, a).
\end{align}
In the spirit of~\cite{dueling_nets}, this condition is enforced by writing $A$ as
\begin{equation}
	A(s, a) = \bar{A}(s, a) - \bar{A}(s, \pi(s)).
\end{equation}
With this parameterization, one can hope to learn an approximation of $A^\pi_\deltat$,
from which a proper hiearchy of actions can be derived for any $\deltat$.

\subsection{Time invariant exploration}
\label{subsec:explo}
To obtain a time invariant reinforcement learning algorithm, a time invariant
exploration scheme is required. All the exploration schemes hereby considered
generate sequences of actions as
\begin{equation}
  \label{eq:explore}
	a^k_\deltat = \pi^\text{explore}_\deltat(s^k_\deltat, z^k_\deltat)
\end{equation}
with $(z^k_\deltat)_k$ a sequence of random variables, independant from the $a$'s and $s$'s.
For instance, $\varepsilon$-greedy exploration naturally falls in this framework, \TODO{[[CT would like to give these details, LB thinks it is not necessary:]] with each $z^k_\deltat$
being a pair composed of a bernoulli random variables $b^k_\deltat$ with probability of being $1$ equal to $\varepsilon$,
and of a uniform categorical random variable $\xi^k_\deltat$ on actions. Exploratory actions are then selected as
\begin{equation}
	a^k_\deltat = (1 - b^k_\deltat) \text{argmax}_{a'} Q^\pi_\deltat(s, a') + b^k_\deltat \xi^k_\deltat.
\end{equation}}

A sufficient condition for the exploratory policy to admit a time continuous
limit is for
$z_\deltat$ to admit a limit, i.e. that the sequence $z_\deltat$ converges to a
well defined continuous stochastic process $z$ as $\deltat$ goes to $0$.
\TODO{I'm convinced that this is true, reviewers might not be.}

The exploration scheme used in~\cite{ddpg} for continuous actions, which consists in adding an
\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions, is naturally time
invariant. It follows Eq. \eqref{eq:explore} with:
\begin{equation}
	\pi^\text{explore}(s_t, z_t) = \pi(s_t) + z_t
\end{equation}
and $z_t$ a time continuous OU process, defined as
\begin{equation}
	dz_t = - z_t \theta dt + \sigma \dbt.
	\label{eq:orn_uhl}
\end{equation}

>>>>

%! TEX root = icml_drau.tex
\section{Reinforcement learning with a continuous time limit}

In what follows, a discrete algorithm with a well defined time continuous
limit is defined. The algorithm relies on three elements, namely, defining and
learning a quantity that contains some information on the ranking of actions
while admitting a time continuous limit, defining exploration methods that
admit a continuous time limit and defining learning rates that induce well
behaved parameter trajectories when $\deltat$ goes to $0$.

As seen in Sec.~\ref{sec:framework}, there is no continuous time limit to
Q-learning. In continuous time, $Q^\pi$ is independent of actions and can thus
not be used to select promising actions.  In near continuous time, $Q^\pi_\deltat$
still depends on actions, and can still be used to choose promising actions
under the current policy. However, when approximating $Q^\pi_\deltat$, if the
approximation error is much bigger than $\bigO(\deltat)$, approximation errors
dominates, and the ranking of actions given by the approximation is likely to
be erroneous.
% When the approximate Q-function is initialized, if the effect of
% actions on the Q-function is order of magnitudes higher than what it should be,
% approximating $Q^\pi_\deltat$ is likely to be difficult.
%Besides, the error is likely
%to be further propagated when the equation used to update our approximate $Q$
%function relies on bootstraping, as is the case for usual temporal difference
%derived methods.

\subsection{Advantage Updating}
\label{subsec:reparam}
To define an object which contains the same information on actions as
$Q^\pi_\deltat$, but admits an action dependant continuous time limit, it is
natural to define
\begin{align}
	A^\pi_\deltat(s, a) &\mathdef \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat},
    \label{eq:adv}
\end{align}
a rescaled version of the advantage function \cite{adv_upd}, as the difference between between
$Q^\pi_\deltat(s, a)$ and $V^\pi\deltat(s)$ is of order $\bigO(\deltat)$.
Contrary to the action state value function, the rescaled advantage function converges when $\deltat$ goes to $0$
to an action dependent quantity which contains the necessary information for policy improvement:

@@

The discretized Q-function can be rewritten as
\begin{equation}
	Q^\pi_\deltat(s, a) = V^\pi_\deltat(s) + \deltat A^\pi_\deltat(s, a).
	\label{eq:reparam_q_pi}
\end{equation}
A natural way of approximating $V^\pi_\deltat$ and $A^\pi_\deltat$ is to apply
\emph{Sarsa} or Q-learning to a reparameterized Q-function approximator
\begin{equation}
	Q_\Theta(s, a) = V_{\theta}(s) + \deltat A_{\psi}(s, a).
\end{equation}
with $\Theta = (\theta, \pi)$. To avoid cumbersome notations, in what follows,
parameter indices are dropped.  At initialization, if both $V_{\theta}$ and
$A_{\psi}$ are initialized independantly of $\deltat$, this parameterization
reasonably scales the contribution of actions relative to states in the value
of $Q$.  Our goal is for $V_\theta$ to approximate $V^\pi_\deltat$ and for
$A_{\psi}$ to approximate $A^\pi_\deltat$.

%\TODO{CT and LB are fighting each other about the importance which should be given to the next paragraph. A impartial judge should give his opinion. (CT is wrong)}
However, the proposed reparameterization does not, on its own, guarantee that
when $Q$ correctly approximates $Q^\pi_\deltat$, $A$ approximates $A^\pi_\deltat$\footnote{
	Ensuring that $A$ do not differ from $A^\pi_\deltat$ by a state dependent function only affects
	$A$'s interpretability. Indeed, for any function $f$, the ranking of actions given by
	$A^\pi_\deltat(s, \cdot)$ and $A^\pi_\deltat(s, \cdot) + f(s)$ is the same.
}.
Indeed, for any given pair $V$, $A$,
\begin{align}
	\tilde{V}(s) &= V(s) - f(s)\\
	\tilde{A}(s, a) &= A(s, a) + \frac{f(s)}{\deltat}
\end{align}
yields the exact same $Q$ function. Consequently, $Q$ can correctly approximate
$Q^\pi$ even if $A$ is arbitrarily far from $A^\pi$.
To enforce identifiability of $A$, one must enforce the consistency equation
\begin{equation}
	Q^\pi_\deltat(s, \pi(s)) - V^\pi_\deltat(s) = 0
\end{equation}
on the approximate $A$ and $V$. This translates to
\begin{equation}
	A(s, \pi(s)) = 0.
\end{equation}
With this additional constraint, if $Q = Q^\pi_\deltat$, then $A = A^\pi_\deltat$ and $V = V^\pi_\deltat$.
\begin{align}
	A^\pi_\deltat(s, a) &= \frac{Q^\pi_\deltat(s,a) - V^\pi_\deltat(s)}{\deltat}\\
		    &= \frac{Q(s,a) - Q(s, \pi(s))}{\deltat}\\
		    &= A(s, a).
\end{align}
In the spirit of~\cite{dueling_nets}, this condition is enforced by writing $A$ as
\begin{equation}
\label{eq:Aparam}
	A(s, a) = \bar{A}(s, a) - \bar{A}(s, \pi(s)).
\end{equation}
With this parameterization, one can hope to learn an approximation of $A^\pi_\deltat$,
from which a proper hiearchy of actions can be derived for any $\deltat$.

\subsection{Time invariant exploration}
\label{subsec:explo}
To obtain a time invariant reinforcement learning algorithm, a time invariant
exploration scheme is required. 
\cite{ddpg} already introduces an exploration scheme for continuous actions, which consists in adding an
\emph{Ornstein-Uhlenbeck}~\cite{orn-uhl} process (OU process) to actions, that is naturally time
discretization invariant. Formally, it is defined as
\begin{equation}
	\pi^\text{explore}(s^k_\deltat, z^k_\deltat) = \pi(s^k_\deltat) + z^k_\deltat
\end{equation}
with $z^k_\deltat$ the discretization of a time continuous OU process, defined as
\begin{equation}
	dz_t = - z_t \theta dt + \sigma \dbt.
	\label{eq:orn_uhl}
\end{equation}
where $B_t$ is a brownian motion. Resulting discretized trajectories converge
to non trivial continuous time trajectories.

This exploration can be extended to schemes of the form
\begin{equation}
  \label{eq:explore}
	a^k_\deltat = \pi^\text{explore}_\deltat(s^k_\deltat, z^k_\deltat)
\end{equation}
with $(z^k_\deltat)_k$ a sequence of random variables, independant from the $a$'s and $s$'s.
A sufficient condition for the exploratory policy to admit a time continuous
limit is for
$z_\deltat$ to admit a limit, i.e. that the sequence $z_\deltat$ converges to a
well defined continuous stochastic process $z$ as $\deltat$ goes to $0$.
% For instance, $\varepsilon$-greedy exploration naturally falls in this framework, \TODO{[[CT would like to give these details, LB thinks it is not necessary:]] with each $z^k_\deltat$
% being a pair composed of a bernoulli random variables $b^k_\deltat$ with probability of being $1$ equal to $\varepsilon$,
% and of a uniform categorical random variable $\xi^k_\deltat$ on actions. Exploratory actions are then selected as
% \begin{equation}
% 	a^k_\deltat = (1 - b^k_\deltat) \text{argmax}_{a'} Q^\pi_\deltat(s, a') + b^k_\deltat \xi^k_\deltat.
% \end{equation}}

% \TODO{I'm convinced that this is true, reviewers might not be.}
